Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2014-07-09 patchy-selection-paper.tex
plos publications format hoops
began simon's edits

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{natbib}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{color}
\usepackage[hidelinks]{hyperref}
%\usepackage{latexml} % for \iflatexml

@@

% short title:  Convergent Evolution on Patchy Landscapes
\title 
{
Convergent Evolution During Local Adaptation to Patchy Landscapes
}

@@

%%%%%%%%%%%
\abstract{
Species often encounter, and adapt to, many patches of locally similar environmental conditions
across their range. 
Such adaptation can occur through convergent evolution as different alleles arise and spread in different patches, 
or through the spread of alleles by migration
acting to synchronize adaptation across the species.
The tension between the two reflects the degree of 
constraint imposed on evolution by the underlying genetic
architecture versus how how effectively selection acts to inhibit the geographic spread of locally
adapted alleles.
This paper studies a model of the balance between these two routes to adaptation in continuous environments with patchy selection pressures.
We address the following questions: 
How long does it take for a new, locally adapted allele to appear in a
patch of habitat where it is favored through new mutation?
Or, through migration from another, already adapted patch?
Which is more likely to occur, as a function of distance between the patches?
How can we tell which has occurred, i.e.\ what population genetic
signal is left by the spread of migrant alleles and how long does this
signal persist for?
To answer these questions we decompose the migration--selection equilibrium
surrounding an already adapted patch
into families of migrant alleles,
in particular treating those rare families that reach new patches
as spatial branching processes.
This provides a way to understand the role of geographic separation between patches 
in promoting convergent adaptation and the genomic signals it leaves behind.
We illustrate these ideas using the convergent evolution of cryptic coloration in the rock pocket mouse, \textit{Chaetodipus intermedius}, as an empirical example.

@@

The convergent evolution of similar phenotypes 
in response to shared selection pressures is a testament to the power of selection 
to repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes
\citep{stern2013genetic,zhen2012parallel,martin2013repeated}. 
Convergent adaptation can also occur within species, if individuals within a species 
adapt to the same environment, in parallel, through different genetic changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes \citep{Arendt:08}.
Such evolution of convergent phenotypes is favored by higher mutational input, 
i.e.\ larger mutational target and/or population size \citep{softsweepsII}.
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
as standing variation for an allele increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
leading to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

@@

We work in a model of continuous geography,
using a variety of known results and new methods.
In section \ref{ss:patchymutation} we develop a simple approximation,
equation \eqref{eqn:mutrate}, for the rate at which patches become established by new mutations. 
The most important conceptual work of the paper occurs in section \ref{ss:patchymigration},
where we develop an understanding of the process by which alleles move from an existing patch 
to colonize a novel patch, culminating in equation \eqref{eqn:migrate} for the rate at which this happens.
We combine these two results in section \ref{ss:probparallel} 
to discuss to the probability of parallel adaptation, equation \eqref{eqn:parallel_prob}.
To understand the genomic signal left by adaptations shared by
migration in section \ref{ss:haplotype_length} we approximate the time
it will take for an allele to transit between patches,
equation \eqref{eqn:mean_tau}, and use this to approximate the length of
haplotype that hitchhikes with it,
equation \eqref{eqn:haplotype_length}. We then quantify how quickly this
haplotype decays within patches equation \eqref{eqn:rec_rate_within} and
use this to obtain properties about the equilibrium distribution of
haplotype lengths, equation 
\eqref{eqn:within_haplotype_length}  in section \ref{ss:patch_haplotype}. 
Finally, in section \ref{ss:applications} we apply this work to understand the geographic scale of
convergent evolution in \textit{Chaetodipus intermedius}.

@@

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{color}
\usepackage{natbib}
\usepackage[hidelinks]{hyperref}

% \usepackage{cite}

% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% turn this on to doublespace and put all figures at the very end.
\newif\ifsubmissionversion
\submissionversionfalse
\newif\ifincludesupplement
\includesupplementfalse

\ifsubmissionversion
\usepackage{endfloat}
\fi

\ifsubmissionversion
% \linespread{1.6}
\doublespacing
\fi

% Use the PLoS provided bibtex style
% OR NOT \bibliographystyle{plos2009}
\bibliographystyle{abbrvnat}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **

@@

% short title:  Convergent Evolution on Patchy Landscapes
\title 
{
Convergent Evolution During Local Adaptation to Patchy Landscapes
}
% Simon: I think the title only partly reflects the scope of the paper (adaptation by migration being underrepresented). What about "Relative importance of migration and convergent evolution for local adaptation in patchy landscapes" or similar?

@@

%%%%%%%%%%%
\abstract{
Species often encounter, and adapt to, many patches of locally similar environmental conditions
across their range. 
Such adaptation can occur through convergent evolution as different alleles arise and spread in different patches, 
or through the spread of alleles by migration
acting to synchronize adaptation across the species.
The tension between the two reflects the degree of 
constraint imposed on evolution by the underlying genetic
architecture versus how effectively selection acts to inhibit the geographic spread of locally
adapted alleles.
This paper studies a model of the balance between these two routes to adaptation in continuous environments with patchy selection pressures.
We address the following questions: 
How long does it take for a novel, locally adapted allele to appear in a
patch of habitat where it is favored through mutation?
Or, through migration from another, already adapted patch?
Which is more likely to occur, as a function of distance between the patches?
How can we tell which has occurred, i.e.\ what population genetic
signal is left by the spread of migrant alleles and how long does this
signal persist for?
To answer these questions we decompose the migration--selection equilibrium
surrounding an already adapted patch
into families of migrant alleles,
in particular treating those rare families that reach new patches
as spatial branching processes.
This provides a way to understand the role of geographic separation between patches 
in promoting convergent adaptation and the genomic signals it leaves behind.
We illustrate these ideas using the convergent evolution of cryptic coloration in the rock pocket mouse, \textit{Chaetodipus intermedius}, as an empirical example.

@@

The convergent evolution of similar phenotypes 
in response to shared selection pressures is a testament to the power of selection 
to repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes
\citep{stern2013genetic,zhen2012parallel,martin2013repeated}. 
Convergent adaptation can also occur within species, if individuals within a species 
adapt to the same environment, in parallel, through different genetic changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes \citep{Arendt:08}.
Such evolution of convergent phenotypes is favored by higher mutational input, 
i.e.\ higher total mutational rate and/or population size \citep{softsweepsII}.
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
as standing variation for an allele increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
leading to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

@@

We work in a model of continuous geography,
using a variety of known results and new methods.
In section \ref{ss:patchymutation} we develop a simple approximation,
equation \eqref{eqn:mutrate}, for the rate at which patches become established by new mutations. 
The most important conceptual work of the paper occurs in section \ref{ss:patchymigration},
where we develop an understanding of the process by which alleles move from an existing patch 
to colonize a novel patch, culminating in equation \eqref{eqn:migrate} for the rate at which this happens.
We combine these two results in section \ref{ss:probparallel} 
to discuss to the probability of parallel adaptation, equation \eqref{eqn:parallel_prob}.
To understand the genomic signal left by adaptations shared by migration, 
in section \ref{ss:haplotype_length}, 
we approximate the time it will take for an allele to transit between patches,
(equation \eqref{eqn:mean_tau}),
and use this to approximate the length of haplotype that hitchhikes with it (equation \eqref{eqn:haplotype_length}). 
We then quantify how quickly this haplotype decays within patches (equation \eqref{eqn:rec_rate_within}) 
and use this to obtain properties about the equilibrium distribution of haplotype lengths 
(equation \eqref{eqn:within_haplotype_length}), in section \ref{ss:patch_haplotype}. 
Finally, in section \ref{ss:applications} we apply this work to understand the geographic scale of
convergent evolution in \textit{Chaetodipus intermedius}.

@@

\ifsubmissionversion
\processdelayedfloats
\fi
